\setlength{\parindent}{0pt}
\setlength{\parskip}{0.6em}

%----------------------------------------------------------------------------
\chapter{Implementation}
\label{chap:archPlan}
%----------------------------------------------------------------------------

\subfile{content/75_archPlan/dsl.tex}

\subfile{content/75_archPlan/arch.tex}

\subfile{content/75_archPlan/process.tex}


\section{Packaging}
how the agent and core containerised
explain the helm chart installing them



\section{Security}
security:
    passwds
        basic auth
    TLS for core server 



\section{Evaluation}

\subsection[Chosen technology]{Chosen technology: Do I recommend Ktor?}

Since I chose Kotlin as the primary technology for my DSL, I needed to select a web server framework that supports Kotlin to easily serve the DSL scripts. Spring Boot would have been the least risky choice, as it is a mature, production ready, and widely used framework with numerous pre-existing libraries. The alternative option was Ktor, a relatively new framework built from the ground up using Kotlin and Coroutines. It claims to be lightweight and flexible. Despite having prior experience with Spring Boot, I wanted to try out Ktor. This thesis was also an experiment with the Ktor framework, and this section reflects my opinion on this new framework.

This experiment was quite exciting, I have learned a lot from it. In my experience, it is relatively easy to create basic things, such as a REST API with authentication and persistent storage. However, it becomes significantly more challenging when you aim to integrate authorization with authentication or when you want to make your code testable. While there is sufficient documentation on the basics, there are no guidelines or best practices for the more advanced use cases, and the examples I mentioned are not even that advanced. Since this is a new framework, the community is small, so there are no forums or blogs on best practices. Many times, I had to figure out on my own how to use the framework effectively.

For SQL persistence, Ktor recommends the Exposed\footnote{\url{https://github.com/JetBrains/Exposed}} ORM library, a lightweight SQL library built on top of JDBC driver for the Kotlin language. Over all, it is a good library, but just like with the core Ktor, figuring out the best practices was challenging. It was particularly difficult when I aimed to make the code testable, as dependency injection was never mentioned in any of the tutorials; instead, global variables were used. I also found that Exposed is not always reliable. When updating Ktor versions, I encountered runtime errors whit previously working code. I had to fix the function responsible for reading the compiled DSL script from the database three times.

For dependency injection (DI), I utilized Koin\footnote{\url{https://insert-koin.io/}}, which is a quite good DI framework. It's unfortunate that it is not emphasized in the core Ktor documentation. It is yet another sign that Ktor needs more time to mature.

In conclusion, I would not recommend using Ktor for enterprise purposes due to the lack of guidelines, reliability issues, and missing features, which make it too risky for business applications. On the other hand, it is easy to learn and use, and enjoyable to work with. Therefore, for hobby projects, it can be a good idea, and perhaps in a couple of years, it will be mature enough for business applications as well.

\subsection[Design Reflections]{Design Reflections: What could have been better}

would have been better: 
        one language for both webhooks and checks
        mounting the compiled script as file instead of downloading

\subsection{Performance}

it is blazing fast
one agent serves only one script, this could be improved, but noproblem, an agent is relatively lightweight

